---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# REDCapExploreR

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

The goal of REDCapExploreR is to provide users with tools to post-process and perform exploratory data analysis on REDCap project data through the REDCap API.

This repository is in early stages active development!

## Installation

You can install the development version of REDCapExploreR like so:

``` r
devtools::install_github("CHOP-CGTInformatics/REDCapExploreR")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(REDCapExploreR)
## basic example code
```

### Recreating a Project's Record Status Dashboard

```{r, eval=FALSE}
redcap_uri <- Sys.getenv("REDCAP_URI")
token <- Sys.getenv("REDCAP_TOKEN")

data <- record_status_dashboard(redcap_uri = redcap_uri, token = token)
```
