---
title: "Getting Started with REDCapExploreR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with REDCapExploreR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(REDCapExploreR)
```

REDCapExploreR provides exploratory tools for REDCap projects accessed through
the REDCap API. Use it to inspect project structure, review form completion,
and identify general data quality findings without modifying the REDCap
project.

## Installation

Install the development version from GitHub and load the package:

```{r install, eval=FALSE}
devtools::install_github("CHOP-CGTInformatics/REDCapExploreR")
library(REDCapExploreR)
```

## REDCap API credentials

Functions that retrieve a REDCap project need an API endpoint and a project
token with appropriate API permissions. Store credentials outside
version-controlled code, such as in environment variables:

```{r credentials, eval=FALSE}
redcap_uri <- Sys.getenv("REDCAP_URI")
token <- Sys.getenv("REDCAP_TOKEN")
```

The examples below assume `redcap_uri` and `token` are set. The package also
includes synthetic objects for credential-free exploration.

## Choose a workflow

The primary workflows are independent. Start with the function that matches
your review goal.

| Goal | Start with |
|---|---|
| Understand project fields and structure | `build_codebook()` |
| Review form completion | `build_record_status_data()` |
| Review data quality findings | `build_quality_report()` |
| Analyze the underlying API tables directly | `pull_redcap_project()` |

## Codebook

`build_codebook()` returns a structured view of project fields, choices, forms,
events, repeating configuration, and project-level metadata.

```{r codebook-api, eval=FALSE}
codebook <- build_codebook(
  redcap_uri = redcap_uri,
  token = token
)

codebook
codebook$fields
```

Use `mock_codebook` to inspect the same output structure without API
credentials.

```{r codebook-mock}
mock_codebook
head(mock_codebook$fields)
```

`view_codebook()` presents each available section as an interactive HTML table:

```{r view-codebook, eval=FALSE}
view_codebook(codebook)
```

See the [`build_codebook()` reference](../reference/build_codebook.html) for
output details and the available viewing options.

To save a codebook output to HTML, simply use:

```{r save-codebook, eval=FALSE}
save_codebook(codebook, file = "path.html")
```

## Record status dashboard

`build_record_status_data()` summarizes REDCap form completion by record.
`plot_record_status()` converts that table into a ggplot heat map similar to the
REDCap Record Status Dashboard.

```{r record-status-api, eval=FALSE}
status_data <- build_record_status_data(
  redcap_uri = redcap_uri,
  token = token
)

plot_record_status(status_data)
```

The synthetic `mock_record_status_data` can be plotted directly:

```{r record-status-mock, fig.alt="Synthetic REDCap record status dashboard heat map."}
plot_record_status(mock_record_status_data)
```

See the [`plot_record_status()` reference](../reference/plot_record_status.html)
for compact mode and display customization.

## Quality report

`build_quality_report()` retrieves records and project structure, runs general
data quality checks, and returns findings, summaries, and standardized metadata.

```{r quality-report-api, eval=FALSE}
report <- build_quality_report(
  redcap_uri = redcap_uri,
  token = token
)

report
report$summaries$project
report$findings
```

Use `mock_quality_report` to explore the report structure and example findings:

```{r quality-report-mock}
mock_quality_report

mock_quality_report$findings |>
  dplyr::select(
    finding_id,
    check,
    issue,
    severity,
    record_id,
    field_name
  )
```

See the [quality report article](quality-report.html) for check definitions,
output structure, interpretation guidance, and core assumptions.

## Advanced project data access

`pull_redcap_project()` retrieves the record and structural metadata tables used
by the quality-report workflow. Use it when a custom analysis needs the
underlying API responses directly.

```{r pull-project-api, eval=FALSE}
project <- pull_redcap_project(
  redcap_uri = redcap_uri,
  token = token
)

names(project)
```

`mock_redcap_project` has the same top-level structure:

```{r pull-project-mock}
names(mock_redcap_project)
```

## Next steps

- Use the [quality report article](quality-report.html) for a deeper review of
  quality checks and report interpretation.
- Browse the [function reference](../reference/index.html) for complete
  arguments and return values.
- Explore `mock_redcap_project`, `mock_record_status_data`, `mock_codebook`, and
  `mock_quality_report` when REDCap credentials are not available.
