---
title: "Getting Started with REDCapExploreR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with REDCapExploreR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(REDCapExploreR)
```

REDCapExploreR provides exploratory tools for REDCap projects accessed through
the REDCap API. The package focuses on project structure, record status
summaries, codebook review, and general data quality checks that help data
managers inspect REDCap data in R.

## Installation

Install the development version from GitHub:

```{r install, eval=FALSE}
devtools::install_github("CHOP-CGTInformatics/REDCapExploreR")
```

Load the package after installation:

```{r load-package}
library(REDCapExploreR)
```

## REDCap API credentials

Most REDCapExploreR workflows need a REDCap API endpoint and an API token with
access to the project being reviewed. A common setup is to store these values in
environment variables and read them at the start of the analysis:

```{r credentials, eval=FALSE}
redcap_uri <- Sys.getenv("REDCAP_URI")
token <- Sys.getenv("REDCAP_TOKEN")
```

The examples below assume `redcap_uri` and `token` have been set. Avoid storing
API tokens directly in scripts, vignettes, or version-controlled files.

## Record Status Dashboard

Use `build_record_status_data()` to retrieve a plotting-friendly table that
summarizes REDCap form completion status by record. Then pass the result to
`plot_record_status()` to create a ggplot heat map similar to the REDCap Record
Status Dashboard.

```{r record-status-api, eval=FALSE}
status_data <- build_record_status_data(
  redcap_uri = redcap_uri,
  token = token
)

plot_record_status(status_data)
```

The package includes a small synthetic dataset for examples and offline
exploration. `mock_record_status_data` is generated from `mock_redcap_project`
using the same internal record-status logic that supports
`build_record_status_data()`.

```{r record-status-mock, fig.alt="Synthetic REDCap record status dashboard heat map."}
mock_record_status_data

plot_record_status(mock_record_status_data)
```

For larger projects, compact mode reduces x-axis text size, truncates long form
labels, and uses lighter tile borders.

```{r record-status-compact, fig.alt="Compact synthetic REDCap record status dashboard heat map.", eval=FALSE}
plot_record_status(mock_record_status_data, mode = "compact")

plot_record_status(
  mock_record_status_data,
  mode = "compact",
  form_label_max = 20
)
```

Because `plot_record_status()` returns a ggplot object, you can add ggplot2
layers to customize labels and display settings.

```{r record-status-custom, fig.alt="Synthetic REDCap record status dashboard heat map with a custom title."}
plot_record_status(mock_record_status_data) +
  ggplot2::labs(title = "REDCap Record Status")
```

## Codebooks

Use `build_codebook()` to pull REDCap project metadata and return a structured
codebook with fields, choices, forms, events, event-instrument mappings,
repeating instrument settings, and project-level summary metadata.

```{r codebook-api, eval=FALSE}
codebook <- build_codebook(
  redcap_uri = redcap_uri,
  token = token
)

codebook$fields
codebook$choices
codebook$forms
```

Use `mock_codebook` to inspect the structure without API credentials.

```{r codebook-mock}
mock_codebook

head(mock_codebook$fields)
mock_codebook$project
```

Use `view_codebook()` to review the codebook as an interactive HTML object with
one table per available section.

```{r view-codebook, eval=FALSE}
viewer <- view_codebook(codebook)
viewer

htmltools::save_html(viewer, "codebook.html")
```

## Quality reports

Use `build_quality_report()` to retrieve project records and metadata, run
general data quality checks, and return findings, summaries, and interpreted
metadata.

```{r quality-report-api, eval=FALSE}
report <- build_quality_report(
  redcap_uri = redcap_uri,
  token = token
)

report$findings
report$summaries$project
report$summaries$fields
```

Use `mock_quality_report` to see the shape of the report output and example
findings.

```{r quality-report-mock}
mock_quality_report

mock_quality_report$findings |>
  dplyr::select(
    check,
    issue,
    severity,
    record_id,
    form_name,
    field_name,
    message
  )

mock_quality_report$summaries$project
```

The report can be limited to selected check groups when a focused review is
needed.

```{r quality-report-checks, eval=FALSE}
report <- build_quality_report(
  redcap_uri = redcap_uri,
  token = token,
  checks = c("missingness", "metadata")
)
```

## Pulling project data

`pull_redcap_project()` retrieves the raw REDCap records and structural metadata
used by the quality reporting workflow. This can be useful when you need the
same API tables for custom exploration.

```{r pull-project-api, eval=FALSE}
project <- pull_redcap_project(
  redcap_uri = redcap_uri,
  token = token
)

names(project)
```

The same top-level structure is represented by `mock_redcap_project`.

```{r pull-project-mock}
names(mock_redcap_project)

mock_redcap_project$metadata
```
