---
title: "Understanding REDCap Quality Reports"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(REDCapExploreR)
```

`build_quality_report()` retrieves REDCap records and project structure,
applies general data quality checks, and organizes the results for review. It
helps data managers and analysts identify records, fields, and metadata that
warrant attention without modifying the REDCap project.

This article explains how to build, review, and interpret a quality report. All
evaluated examples use `mock_quality_report`, a synthetic report included with
the package, so they do not require REDCap credentials.

## How to Build a Report

### Generate a report object

Store the REDCap API URI and token outside version-controlled code, such as in
environment variables, and pass them to `build_quality_report()`:

```{r build-report, eval=FALSE}
report <- build_quality_report(
  redcap_uri = Sys.getenv("REDCAP_URI"),
  token = Sys.getenv("REDCAP_TOKEN")
)
```

### Choose checks

The default call runs all five check groups: `missingness`, `metadata`,
`outliers`, `operational`, and `consistency`. Use `checks` for a focused review:

```{r focused-report, eval=FALSE}
report <- build_quality_report(
  redcap_uri = Sys.getenv("REDCAP_URI"),
  token = Sys.getenv("REDCAP_TOKEN"),
  checks = c("missingness", "metadata")
)
```

Summaries and standardized metadata are always returned. Project- and
record-level `finding_count` values reflect only the selected check groups.
`mock_quality_report` was built with all default checks and thresholds.

### Set heuristic thresholds

Two arguments control heuristic findings. They do not change the underlying
REDCap data or missing-rate calculations.

| Argument | Default | Finding affected | Lower values | Higher values |
|---|---:|---|---|---|
| `sparse_threshold` | `0.95` | `unexpected_sparse_field` | Flag more fields | Flag fewer fields |
| `outlier_iqr_multiplier` | `3` | `numeric_iqr_outlier` | Flag more values | Flag fewer values |

For each field, `sparse_threshold` is compared with
`missing_count / record_count` across rows eligible for missingness assessment.
A field is flagged when its missing rate is greater than or equal to the
threshold. At the default `0.95`, a field with 19 missing values among 20
applicable rows is flagged.

For each numeric, non-choice field other than the record identifier,
`outlier_iqr_multiplier` defines the heuristic range:

```text
lower bound = Q1 - outlier_iqr_multiplier * IQR
upper bound = Q3 + outlier_iqr_multiplier * IQR
```

Values strictly outside the range are flagged. The default multiplier of `3`
uses wider bounds than the conventional `1.5 * IQR` rule and is less sensitive
to moderate extremes. The check requires at least four observed values and a
nonzero IQR. It does not affect metadata-defined validation ranges or
future-date findings.

## How to Review a Report

A practical review usually follows four steps:

1. Inspect the project overview.
2. Count findings by check, issue, and severity.
3. Filter record-field findings into a review queue.
4. Join field metadata when labels or field types add context.

### Start with the report overview

Printing the report gives a short project-level overview. The object is a list
with `findings`, `summaries`, and `metadata` elements.

```{r report-overview}
mock_quality_report

names(mock_quality_report)
names(mock_quality_report$summaries)
names(mock_quality_report$metadata)
```

Begin with the project summary and a count of findings. This shows project size
and structure alongside the types of review items found.

```{r report-triage}
mock_quality_report$summaries$project

mock_quality_report$findings |>
  dplyr::count(check, issue, severity, sort = TRUE)
```

### Review actionable findings

Use `scope` to separate record-field findings from field-level metadata and
aggregate findings. Event and repeat columns preserve context for longitudinal
and repeating projects.

```{r record-findings}
record_findings <- mock_quality_report$findings |>
  dplyr::filter(scope == "record_field") |>
  dplyr::select(
    finding_id,
    severity,
    issue,
    record_id,
    event_name,
    repeat_instrument,
    repeat_instance,
    field_name,
    value,
    expected
  )

record_findings
```

Filter by record, issue, form, or severity to create a review queue. Severity is
a package-provided prioritization aid, not a clinical or regulatory
classification.

```{r review-one-record}
mock_quality_report$findings |>
  dplyr::filter(record_id == "C002") |>
  dplyr::select(
    finding_id,
    severity,
    issue,
    form_name,
    field_name,
    value,
    expected,
    message
  )
```

Join findings to field metadata when labels and field types are useful during
review:

```{r enrich-findings}
field_reference <- mock_quality_report$metadata$fields |>
  dplyr::select(field_name, field_label, field_type, required_field)

mock_quality_report$findings |>
  dplyr::filter(!is.na(field_name)) |>
  dplyr::inner_join(field_reference, by = "field_name") |>
  dplyr::select(
    finding_id,
    issue,
    record_id,
    field_name,
    field_label,
    field_type,
    required_field
  )
```

## Quality Report Structure

### Findings

`findings` is a standardized tibble with one row per detected issue. A report
with no findings returns an empty tibble with the same columns.

| Column | Meaning |
|---|---|
| `finding_id` | Sequential identifier assigned within the report. |
| `check` | Check group that produced the finding. |
| `issue` | Stable issue name suitable for filtering and counting. |
| `severity` | Package-assigned `warning` or `info` review priority. |
| `scope` | `record_field` for record-field findings or `field` for metadata and aggregate findings. |
| `record_id` | REDCap record identifier when the finding is record-specific. |
| `form_name` | REDCap instrument associated with the finding. |
| `event_name` | Unique REDCap event name when available. |
| `repeat_instrument` | Repeating instrument name when available. |
| `repeat_instance` | Repeating instance number when available. |
| `field_name` | REDCap field associated with the finding. |
| `value` | Observed or aggregate value that triggered the finding. |
| `expected` | Concise description of the expected state or range. |
| `message` | Human-readable explanation of the finding. |

The current findings are grouped below by check.

#### Missingness

These checks describe absent data only where a field is eligible for
assessment. They identify required values that are missing on specific rows and
fields whose eligible values are almost entirely missing.

| Issue | What it identifies | Severity and scope |
|---|---|---|
| `required_field_missing` | A required field is blank where form status, project structure, and supported branching logic permit assessment. | Warning, `record_field` |
| `unexpected_sparse_field` | A field's missing rate meets or exceeds `sparse_threshold` among applicable rows. | Info, `field` |

#### Metadata

Metadata checks inspect project design rather than the validity of individual
record values. They highlight labels, choice definitions, and free-text fields
that may need review by a project designer or data manager.

| Issue | What it identifies | Severity and scope |
|---|---|---|
| `duplicate_field_label` | The same nonblank label is used by multiple field names. | Info, `field` |
| `missing_field_label` | A field has no label. | Info, `field` |
| `missing_choice_definition` | A radio, dropdown, or checkbox field has no explicit choice definition. | Warning, `field` |
| `high_risk_free_text` | A text or notes field name or label contains terms associated with notes, comments, other details, or similar free text. | Info, `field` |

Yes/No and True/False fields use implicit REDCap choices, so they do not require
explicit choice definitions.

#### Outliers

Outlier checks combine rules configured in REDCap with general review
heuristics. They identify values outside metadata bounds, unusually distant
numeric values, and dates after the report is built.

| Issue | What it identifies | Severity and scope |
|---|---|---|
| `outside_validation_range` | A valid numeric, date, datetime, or time value is outside `text_validation_min` or `text_validation_max`. | Warning, `record_field` |
| `numeric_iqr_outlier` | A numeric, non-choice value is outside the configured IQR heuristic. | Info, `record_field` |
| `future_date` | A valid date or datetime has a calendar date after the system date captured when the report is built. | Info, `record_field` |

Choice fields are excluded from IQR outlier detection even when their stored
codes are numeric, because those codes represent categories rather than
quantities.

#### Operational

Operational checks review REDCap workflow state rather than field values. The
current check identifies applicable forms that have not been marked Complete.

| Issue | What it identifies | Severity and scope |
|---|---|---|
| `incomplete_form_status` | An applicable form status is missing or is not Complete. | Info, `record_field` |

#### Consistency

Consistency checks compare stored values with their field definitions and
identify contradictory checkbox selections. These findings are tied to either
a specific record-field value or an aggregate checkbox pattern.

| Issue | What it identifies | Severity and scope |
|---|---|---|
| `invalid_choice_value` | A raw radio, dropdown, Yes/No, or True/False code is absent from the field's choices. | Warning, `record_field` |
| `invalid_validation_format` | A text value does not satisfy its configured REDCap validation. | Warning, `record_field` |
| `checkbox_no_values_selected` | No options are selected across assessed rows with explicit checkbox exports. | Info, `field` |
| `checkbox_none_with_other` | A none-like option is selected with another checkbox option. | Warning, `record_field` |

### Summaries

`summaries` provides five views of the normalized project and report findings.

#### Project

`summaries$project` is a one-row overview:

- `record_count`: number of distinct record identifiers.
- `raw_row_count`: number of rows returned by the records API, including event
  and repeating rows.
- `field_count`, `form_count`, and `event_count`: project structure counts.
- `repeating_enabled`: whether repeating configuration was returned by REDCap.
- `missing_rate`: weighted missing fraction across applicable field-row cells.
- `finding_count`: total findings from the selected check groups.

```{r project-summary}
mock_quality_report$summaries$project
```

#### Forms and fields

`summaries$forms` reports field counts, required-field counts, and weighted
missing rates by form. `summaries$fields` provides the detailed missingness
profile:

- `record_count` is the number of applicable rows assessed for the field. It is
  not necessarily the number of distinct records because events and repeating
  instruments can contribute multiple rows per record.
- `missing_count` and `observed_count` partition applicable rows.
- `missing_rate` is `missing_count / record_count`, or `NA` when no rows are
  applicable.
- `distinct_count` counts unique nonmissing values.

```{r field-summaries}
mock_quality_report$summaries$forms

mock_quality_report$summaries$fields |>
  dplyr::arrange(dplyr::desc(missing_rate), field_name) |>
  dplyr::select(
    field_name,
    form_name,
    required_field,
    record_count,
    missing_count,
    missing_rate,
    distinct_count
  )
```

Project and form missing rates are weighted across applicable field-row cells;
they are not simple averages of field-level percentages.

#### Records and events

`summaries$records` contains one row per record. `missing_field_count` sums
missing applicable field values across that record's events and repeat
instances. `finding_count` counts record-specific findings; field-level findings
do not belong to an individual record.

`summaries$events` counts raw API rows by unique event name. It is empty for
nonlongitudinal projects.

```{r record-event-summaries}
mock_quality_report$summaries$records |>
  dplyr::arrange(dplyr::desc(finding_count), record_id)

mock_quality_report$summaries$events
```

### Metadata

`metadata` preserves the standardized project structure used by the checks.

| Element | Contents |
|---|---|
| `fields` | Standardized REDCap field metadata used by the report. |
| `forms` | One row per form with field, required-field, and choice-field counts. |
| `events` | Standardized longitudinal event metadata. |
| `event_instruments` | Standardized event-to-form mapping. |
| `instruments` | Instrument names and labels returned by REDCap. |
| `repeating_instruments` | Repeating event and instrument configuration returned by REDCap. |
| `choices` | Parsed explicit choices and implicit Yes/No and True/False choices. |
| `validation` | Fields with validation types or minimum/maximum bounds. |
| `branching` | Fields with branching logic and parsed referenced field names. |
| `record_id_field` | Field name used as the project record identifier. |
| `source` | Source of the normalized project; API reports use `"api"`. |

These tables support project-level review and help interpret findings:

```{r metadata-examples}
mock_quality_report$metadata$repeating_instruments

mock_quality_report$metadata$validation

mock_quality_report$metadata$branching
```

## How Checks Work

Different check groups intentionally evaluate different sets of rows. In this
table, **form status** determines whether the row can enter the check,
**branching logic** determines whether a currently hidden field is excluded,
and **blank values** describes what happens after a row and field are eligible.

| Evaluation | Form status | Branching logic | Blank values |
|---|---|---|---|
| Missingness | Only Unverified or Complete forms are evaluated. | A field is evaluated only where supported logic indicates it is shown. | A blank eligible value is counted as missing. |
| Choice and text validation | Stored values are evaluated for any form status, including Incomplete. | A stored value is evaluated even if branching currently hides the field. | Blank values do not produce choice or format findings. Eligible blanks are handled separately by missingness checks. |
| Checkbox consistency | Only Unverified or Complete forms are evaluated. | A checkbox is evaluated only where supported logic indicates it is shown. | Explicit `0` values are meaningful; rows where every option is blank are skipped. |
| Metadata | Record rows and form status are not used. | Branching logic does not control metadata checks. | A blank label or choice definition can trigger a metadata finding; blank record values are not examined. |

"Supported logic" refers to the subset of REDCap branching expressions the
package can evaluate, described under Core Assumptions. Event-to-form mappings
and repeating configuration are respected whenever a check evaluates record
data. For example, a blank field on a Complete form counts as missing only when
supported branching logic shows that field. By contrast, a nonblank malformed
value on an Incomplete form can still produce a validation finding, even when
branching currently hides the field.

### Missingness

Missingness is evaluated in two stages. The report first determines whether a
field should be assessed on an exported row, then determines whether the field
has an observed value. This prevents structural blanks from events, repeating
instruments, branching logic, and unfinished forms from being treated as
missing data.

#### Form status

Raw REDCap form status determines whether a row contributes to missingness:

| Form status | Included in missingness? | `incomplete_form_status`? |
|---|---:|---:|
| Incomplete (`0`) | No | Yes |
| Unverified (`1`) | Yes | Yes |
| Complete (`2`) | Yes | No |
| Blank value | No | Yes, when the form is otherwise applicable |
| Status column unavailable | No | No status finding can be generated |

If a form has no exported `<form_name>_complete` column, its fields have no
assessable rows and their missing rates are `NA`.

#### Field applicability

Among rows with an assessable status, applicability accounts for:

1. Event-to-form mappings in longitudinal projects.
2. Repeating event and instrument configuration.
3. The repeating instrument associated with each exported row.
4. Supported branching logic that determines whether the field is shown.

For applicable non-checkbox fields, `NA` and blank or whitespace-only values
are missing. A missing required field produces `required_field_missing`. Any
required or optional field can produce `unexpected_sparse_field` when its
missing rate meets `sparse_threshold`.

#### Checkbox behavior

REDCap exports one `0`/`1` column per checkbox choice. The report evaluates
missingness at the parent-field level:

- One or more checked (`1`) choices are observed.
- All choices explicitly unchecked (`0`) are also observed. This is a
  meaningful no-selection state.
- The parent field is missing only when every exported choice value is `NA` or
  blank on an otherwise applicable row.

On an Incomplete form, checkbox exports are ignored for missingness. On an
Unverified or Complete form, an all-zero state is observed, while an all-missing
set can contribute to sparse-field and required-field findings.

Checkbox consistency findings do not redefine the all-zero state as missing:

- `checkbox_no_values_selected` is one field-level prompt when no option is
  selected across assessed rows with explicit checkbox exports.
- `checkbox_none_with_other` is a record-field warning when an option labeled
  "None", "No", "Not applicable", "N/A", "NA", or "None of the above" is
  selected with another option.

### Metadata-defined validation

REDCap metadata records the validation selected by a project designer in the
REDCap field configuration UI. For text fields, the selected validation type
and optional minimum or maximum are exported in the `text_validation_*`
columns. Choice definitions configured in REDCap are similarly exported as
field metadata.

The report compares nonblank raw values with these user-configured rules. It
inspects structurally applicable rows regardless of whether a form is
Incomplete, Unverified, or Complete. Branching logic does not suppress a stored
value that violates its metadata rule.

#### Choice values

`invalid_choice_value` identifies raw codes absent from metadata choices for
radio, dropdown, Yes/No, and True/False fields. Yes/No and True/False use their
implicit REDCap `0`/`1` definitions.

Checkbox fields are excluded. Their metadata codes identify separate option
columns, while each exported cell contains a `0` or `1` selection state.

#### Why the raw export is used

REDCap choice fields have a stored code and a displayed label. Validation must
compare the stored code with the codes defined in metadata, so the report
requests raw values rather than labels. For example, a dropdown choice stored
as code `1` remains `"1"` instead of being replaced by its display label.

Automatic column type guessing is also disabled. This prevents a column of `0`
and `1` codes from being converted to logical `FALSE` and `TRUE`, which would no
longer match the REDCap choice definitions. Empty API cells are imported as
missing, while a literal code such as `"NA"` is preserved as an actual value.

Fields without explicit radio, dropdown, or checkbox definitions produce
`missing_choice_definition` instead of an invalid-choice finding.

#### Text formats and bounds

`invalid_validation_format` applies only to `text` fields with a supported
`text_validation_type_or_show_slider_number`. The report does not infer
validation from R classes, field names, labels, or observed values.

| Family | Metadata examples | Validation behavior |
|---|---|---|
| Integer | `integer` | Integer syntax; scientific notation is rejected. |
| Number | `number`, `number_1dp` through `number_4dp`, comma-decimal variants | Configured decimal mark and precision; a leading zero is optional. |
| Date | `date_dmy`, `date_mdy`, `date_ymd` | Canonical API value `YYYY-MM-DD`; impossible dates are rejected. |
| Datetime | DMY, MDY, and YMD `datetime_*` types | Canonical YMD export with minutes or seconds, according to the validation type. |
| Time | `time`, `time_hh_mm_ss`, `time_mm_ss` | Configured format and clock boundaries. |
| Email | `email` | Structural email format, not deliverability. |
| Phone | `phone`, `phone_australia` | Common punctuation and regional digit structure. |

The DMY, MDY, and YMD suffix controls REDCap's data-entry display. API exports
still use canonical YMD date and datetime values. Number validations accept
values with or without a leading zero, such as `0.850` and `.850`.

When provided, `text_validation_min` and `text_validation_max` produce
`outside_validation_range` for valid numeric, date, datetime, and time values
outside the configured bounds. Either bound can be used alone. Malformed values
produce `invalid_validation_format` but not range or future-date findings.

Use issue names to review metadata-defined validation findings together. The
mock report contains a range violation but no malformed formats or invalid
choice codes.

```{r validation-findings}
validation_issues <- c(
  "invalid_choice_value",
  "invalid_validation_format",
  "outside_validation_range"
)

mock_quality_report$findings |>
  dplyr::filter(issue %in% validation_issues) |>
  dplyr::select(record_id, field_name, issue, value, expected)
```

### Distribution and future-date checks

`numeric_iqr_outlier` is a distribution-based heuristic, not a REDCap metadata
rule. Its behavior depends on sample size, field distribution, and
`outlier_iqr_multiplier`.

`future_date` applies to valid date and datetime fields. It compares the
exported calendar date with `Sys.Date()` captured when the report runs. A date
after that system date is flagged; a datetime later on the same calendar day is
not. Malformed temporal values are handled by `invalid_validation_format`
instead.

## Core Assumptions

### Findings are review prompts

The checks identify items worth reviewing. They do not establish clinical
correctness, protocol compliance, or regulatory compliance. An `info` finding
can still be important, and a report with no findings does not prove that the
data are correct.

`finding_id` is assigned when a report is built and is not persistent across
runs. Use the issue and REDCap context columns when tracking findings
externally.

### Branching logic support is intentionally bounded

The branching evaluator supports ordinary field comparisons, checkbox choice
references, parentheses, and `and`/`or` combinations. A valid expression that
evaluates to `NA` is treated as not shown for that row.

REDCap functions, smart variables, and expressions outside the supported subset
are treated as applicable rather than used to hide a potentially required
value. The report does not produce findings for apparently missing branching
references because event-qualified fields, cross-form fields, and smart
variables cannot be classified reliably from bracketed tokens alone.

### Counts reflect REDCap's row structure

A REDCap record can produce multiple API rows because of events and repeating
instances. Project `record_count` counts distinct records, `raw_row_count`
counts exported rows, field `record_count` counts applicable rows, and event
`row_count` counts exported rows per event. Choose the measure that matches the
question being asked.

### Results reflect the current project export

Descriptive fields are removed before analysis because they do not store record
values. Structural metadata retrieval failures are reported as errors rather
than silently converted to empty project structure. Results reflect the records
and metadata available when the report is built, including the token's project
access and export configuration.
