---
title: "Understanding REDCap Quality Reports"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(REDCapExploreR)
```

`build_quality_report()` retrieves REDCap records and project structure, applies
general data quality checks, and organizes the results for review. The report is
intended to help data managers and analysts identify records, fields, and
metadata that warrant attention. It does not modify the REDCap project.

This article explains how to build and navigate a report. All evaluated
examples use `mock_quality_report`, a synthetic report included with the
package, so they do not require REDCap credentials.

## How to Build a Report

### Generate a report object

Store the REDCap API URI and token outside version-controlled code, such as in
environment variables, and pass them to `build_quality_report()`:

```{r build-report, eval=FALSE}
report <- build_quality_report(
  redcap_uri = Sys.getenv("REDCAP_URI"),
  token = Sys.getenv("REDCAP_TOKEN")
)
```

`checks` controls which findings are generated. Summaries and interpreted
metadata are still returned, while `finding_count` values in the project and
record summaries reflect the selected check groups. `mock_quality_report` was
built with all default check groups and thresholds.

The default call runs all five check groups: `missingness`, `metadata`,
`outliers`, `operational`, and `consistency`. Use `checks` to run a focused
review:

```{r focused-report, eval=FALSE}
report <- build_quality_report(
  redcap_uri = Sys.getenv("REDCAP_URI"),
  token = Sys.getenv("REDCAP_TOKEN"),
  checks = c("missingness", "metadata"),
  progress = "none"
)
```

### Set heuristic thresholds

Two arguments control when heuristic findings are produced. They do not change
the underlying REDCap data or missing-rate calculations, but they can change
findings and the associated `finding_count` summaries.

`sparse_threshold` is a proportion greater than `0` and less than or equal to
`1`. For each field, the report calculates
`missing_count / record_count` across rows eligible for missingness assessment.
An `unexpected_sparse_field` finding is generated when that missing rate is
greater than or equal to `sparse_threshold`. The default `0.95` therefore flags
a field when at least 95% of its assessed values are missing. For example, a
field with 19 missing values among 20 applicable rows has a missing rate of
`0.95` and is flagged at the default threshold. Raising the threshold produces
fewer sparse-field findings; lowering it makes the check more sensitive.

`outlier_iqr_multiplier` is a positive number used only by the
`numeric_iqr_outlier` check. For each exported numeric, non-choice field other
than the record identifier, the report calculates the first quartile (Q1),
third quartile (Q3), and
`IQR = Q3 - Q1`, then defines the heuristic range as:

```text
lower bound = Q1 - outlier_iqr_multiplier * IQR
upper bound = Q3 + outlier_iqr_multiplier * IQR
```

Values strictly outside that range are flagged. The default multiplier of `3`
uses wider bounds than the conventional `1.5 * IQR` rule and is therefore less
sensitive to moderate extremes. Smaller multipliers produce more findings;
larger multipliers produce fewer. The check requires at least four observed
values and a nonzero IQR, excludes choice fields and the record identifier, and
does not affect REDCap validation-range or future-date findings.

### Control progress output

`progress = "auto"` is the default. It enables normally throttled progress
updates in interactive sessions, such as an R console, and suppresses them in
noninteractive scripts and most automated jobs. `progress = "show"` enables
progress in all sessions and forces every report step to render. Use
`progress = "none"` when no progress output should be emitted.

### Start with the report overview

Printing the report gives a short project-level overview. The object itself is
a list with `findings`, `summaries`, and `metadata` elements.

```{r report-overview}
mock_quality_report

names(mock_quality_report)
names(mock_quality_report$summaries)
names(mock_quality_report$metadata)
```

Begin with the project summary and a count of findings by check and issue. This
shows the size and structure of the project alongside the types of review items
that were found.

```{r report-triage}
mock_quality_report$summaries$project

mock_quality_report$findings |>
  dplyr::count(check, issue, severity, sort = TRUE)
```

### Review actionable findings

Use `scope` to distinguish findings tied to a specific record and field from
field-level metadata findings. The context columns remain available for
longitudinal and repeating projects.

```{r record-findings}
record_findings <- mock_quality_report$findings |>
  dplyr::filter(scope == "record_field") |>
  dplyr::select(
    finding_id,
    severity,
    issue,
    record_id,
    event_name,
    repeat_instrument,
    repeat_instance,
    field_name,
    value,
    expected
  )

record_findings
```

Filter by record, issue, form, or severity to create a review queue. Severity is
a package-provided prioritization aid, not a clinical or regulatory
classification.

```{r review-one-record}
mock_quality_report$findings |>
  dplyr::filter(record_id == "C002") |>
  dplyr::select(
    finding_id,
    severity,
    issue,
    form_name,
    field_name,
    value,
    expected,
    message
  )
```

Join findings to interpreted field metadata when labels and field types are
useful during review:

```{r enrich-findings}
field_reference <- mock_quality_report$metadata$fields |>
  dplyr::select(field_name, field_label, field_type, required_field)

mock_quality_report$findings |>
  dplyr::filter(!is.na(field_name)) |>
  dplyr::inner_join(field_reference, by = "field_name") |>
  dplyr::select(
    finding_id,
    issue,
    record_id,
    field_name,
    field_label,
    field_type,
    required_field
  )
```

## Missingness

Missingness is evaluated in two stages. The report first determines whether a
field should be assessed on a particular exported row. It then determines
whether that field has an observed value. This prevents structural blanks from
events, repeating instruments, branching logic, and unfinished forms from
being treated as missing data.

### Form status determines which rows are assessed

A field is eligible for missingness assessment only when its raw REDCap form
status column, `<form_name>_complete`, is Unverified (`1`) or Complete (`2`).
Rows with an Incomplete (`0`), blank, or unavailable form status do not
contribute to field `record_count`, `missing_count`, or `missing_rate`.

Non-Complete statuses, including Incomplete, Unverified, and blank values, are
still reviewable through the `incomplete_form_status` operational finding. The
distinction is intentional: Unverified forms are treated as sufficiently
assessable but have not been marked Complete, while Incomplete forms can contain
expected blanks that should not inflate missingness findings. If a form has no
exported completion status column, its fields have no assessable rows and their
missing rates are `NA`.

### Per-field missingness evaluation

Among rows with an assessable form status, applicability also accounts for:

1. The event-to-form mapping in longitudinal projects.
2. Repeating event and instrument configuration.
3. The repeating instrument associated with each exported row.
4. Supported branching logic that determines whether the field is shown.

For applicable rows, `NA` and blank or whitespace-only values are missing.
Each field summary reports:

- `record_count`: applicable rows assessed for the field.
- `missing_count`: applicable rows with a missing value.
- `observed_count`: applicable rows with an observed value.
- `missing_rate`: `missing_count / record_count`, or `NA` when no rows are
  applicable.

A missing required field produces a record-level `required_field_missing`
finding. Any field, whether required or optional, can produce a field-level
`unexpected_sparse_field` finding when its missing rate meets
`sparse_threshold`.

### Checkbox fields

REDCap exports a checkbox field as one `0`/`1` column per choice. The report
evaluates missingness at the parent-field level rather than treating every
unchecked choice as a missing value:

- One or more checked (`1`) choices make the parent checkbox field observed.
- All choices explicitly unchecked (`0`) are also observed. This represents a
  meaningful no-selection state rather than a missing response.
- The parent field is missing only when every exported choice value is `NA` or
  blank on an otherwise applicable row.
- A checked explicit option labeled "None", "No", "Not applicable", "N/A",
  "NA", or "None of the above" is an observed response. Selecting one of these
  with another option can separately produce a `checkbox_none_with_other`
  consistency finding.

The form-status rule is applied first. Consequently, checkbox exports on an
Incomplete form are ignored for missingness. On an Unverified or Complete form,
an explicit all-zero state does not contribute to missingness, while an
all-missing set of choice exports can contribute to sparse-field summaries and,
for a required checkbox field, a `required_field_missing` finding.

Both checkbox consistency findings use only applicable rows from Unverified or
Complete forms. When no options are selected anywhere for a checkbox field, the
report produces one field-level `checkbox_no_values_selected` informational
finding. It is a prompt to confirm that the no-selection state is intentional,
not a missingness finding.

## Quality Report Structure

### Findings

`findings` is a standardized tibble with one row per detected issue. A report
with no findings returns an empty tibble with the same columns.

| Column | Meaning |
|---|---|
| `finding_id` | Sequential identifier assigned within the report. |
| `check` | Check group that produced the finding. |
| `issue` | Stable issue name suitable for filtering and counting. |
| `severity` | Package-assigned `warning` or `info` review priority. |
| `scope` | `record_field` for record-level findings or `field` for metadata and aggregate findings. |
| `record_id` | REDCap record identifier when the finding is record-specific. |
| `form_name` | REDCap instrument associated with the finding. |
| `event_name` | Unique REDCap event name when available. |
| `repeat_instrument` | Repeating instrument name when available. |
| `repeat_instance` | Repeating instance number when available. |
| `field_name` | REDCap field associated with the finding. |
| `value` | Observed value or aggregate value that triggered the finding. |
| `expected` | Concise description of the expected state or range. |
| `message` | Human-readable explanation of the finding. |

The following checks are currently performed.

| Check | Issue | What it identifies | Severity and scope |
|---|---|---|---|
| Missingness | `required_field_missing` | A required field is blank on a row where the form and branching logic are applicable and completion status permits missingness assessment. | Warning, `record_field` |
| Missingness | `unexpected_sparse_field` | A field's missing rate is at or above `sparse_threshold` among its applicable rows. | Info, field |
| Metadata | `duplicate_field_label` | The same nonblank field label is used by multiple field names. | Info, field |
| Metadata | `missing_field_label` | A field has no label. | Info, field |
| Metadata | `missing_choice_definition` | A radio, dropdown, or checkbox field has no explicit choice definition. | Warning, field |
| Metadata | `orphaned_branching_reference` | Branching logic references a field not present in project metadata. | Warning, field |
| Metadata | `high_risk_free_text` | A text or notes field name or label contains terms associated with notes, comments, other details, or similar free text. | Info, field |
| Outliers | `outside_validation_range` | A numeric value is below or above a REDCap metadata validation bound. | Warning, `record_field` |
| Outliers | `numeric_iqr_outlier` | A numeric, non-choice field falls outside the configured IQR heuristic. At least four observed values and a nonzero IQR are required. | Info, `record_field` |
| Outliers | `future_date` | A field with date validation contains a `YYYY-MM-DD` value after the report date. | Info, `record_field` |
| Operational | `incomplete_form_status` | An applicable form status is missing or is not Complete. | Info, `record_field` |
| Consistency | `checkbox_no_values_selected` | No options are selected for a checkbox field across applicable Unverified or Complete forms. | Info, field |
| Consistency | `checkbox_none_with_other` | A checkbox option labeled None, No, Not applicable, N/A, NA, or None of the above is selected with another option. | Warning, `record_field` |

Choice fields are excluded from IQR outlier detection even when their codes are
numeric. Yes/No and True/False fields have implicit REDCap choices, so they do
not require an explicit choice definition.

### Summaries

`summaries` provides five views of the normalized project and report findings.

#### Project

`summaries$project` is a one-row overview:

- `record_count`: number of distinct record identifiers.
- `raw_row_count`: number of rows returned by the records API, including event
  and repeating rows.
- `field_count`, `form_count`, and `event_count`: project structure counts.
- `repeating_enabled`: whether repeating configuration was returned by REDCap.
- `missing_rate`: weighted missing fraction across applicable field-row cells.
- `finding_count`: total number of findings from the selected check groups.

```{r project-summary}
mock_quality_report$summaries$project
```

#### Forms and fields

`summaries$forms` reports field counts, required-field counts, and weighted
missing rates by form. `summaries$fields` gives the most detailed missingness
profile:

- `record_count` is the number of applicable rows assessed for the field. It is
  not necessarily the number of distinct records because longitudinal and
  repeating projects can contribute multiple rows per record.
- `missing_count` and `observed_count` partition applicable rows.
- `missing_rate` is `missing_count / record_count`, or `NA` when no rows are
  applicable.
- `distinct_count` counts unique nonmissing values.

```{r field-summaries}
mock_quality_report$summaries$forms

mock_quality_report$summaries$fields |>
  dplyr::arrange(dplyr::desc(missing_rate), field_name) |>
  dplyr::select(
    field_name,
    form_name,
    required_field,
    record_count,
    missing_count,
    missing_rate,
    distinct_count
  )
```

Project and form missing rates are weighted across applicable field-row cells.
They are not simple averages of field-level percentages.

#### Records and events

`summaries$records` contains one row per record. `missing_field_count` sums
missing applicable field values across that record's events and repeat
instances, while `finding_count` counts record-specific findings. Field-level
metadata findings do not belong to an individual record.

`summaries$events` counts raw API rows by unique event name. It is empty for
nonlongitudinal projects.

```{r record-event-summaries}
mock_quality_report$summaries$records |>
  dplyr::arrange(dplyr::desc(finding_count), record_id)

mock_quality_report$summaries$events
```

### Metadata

`metadata` preserves and interprets project structure used by the checks.

| Element | Contents |
|---|---|
| `fields` | Standardized REDCap field metadata used by the report. |
| `forms` | One row per form with field, required-field, and choice-field counts. |
| `events` | Standardized longitudinal event metadata. |
| `event_instruments` | Standardized event-to-form mapping. |
| `instruments` | Instrument names and labels returned by REDCap. |
| `repeating_instruments` | Repeating event and instrument configuration returned by REDCap, with normalized join columns. |
| `choices` | Parsed explicit choices and implicit Yes/No and True/False choices. |
| `validation` | Fields with validation types or minimum/maximum bounds. |
| `branching` | Fields with branching logic and parsed referenced field names. |
| `record_id_field` | Field name used as the project record identifier. |
| `source` | Source of the normalized project; API reports use `"api"`. |

These tables are useful for interpreting findings and for project-level review
that is not represented by a finding. For example, inspect repeating structure
and validation rules directly:

```{r metadata-examples}
mock_quality_report$metadata$repeating_instruments

mock_quality_report$metadata$validation

mock_quality_report$metadata$branching
```

## Core Assumptions

### Findings are review prompts

The checks are broad heuristics intended to identify items worth reviewing.
They do not establish clinical correctness, protocol compliance, or regulatory
compliance. An `info` finding can still be important, and a report with no
findings does not prove that the data are correct.

`finding_id` is assigned when a report is built and should not be treated as a
persistent identifier across runs. Use issue and REDCap context columns when
tracking findings externally.

### Branching logic support is intentionally bounded

The branching evaluator supports ordinary field comparisons, checkbox choice
references, parentheses, and `and`/`or` combinations. A valid expression that
evaluates to `NA` is treated as not shown for that row.

REDCap functions, smart variables, and other expressions outside the supported
subset are treated as applicable rather than used to hide a potentially
required value. The metadata check separately reports references to fields that
are absent from project metadata. Complex branching findings should therefore
be reviewed against REDCap itself.

### Outlier checks depend on metadata and sample size

Validation-range findings depend on numeric minimum and maximum values defined
in REDCap metadata. IQR findings are distribution-based and require at least
four observed values and a nonzero IQR. Their usefulness depends on the field's
distribution and the selected multiplier.

Future-date checks apply only to fields whose REDCap validation type contains
`date`, and values are interpreted in `YYYY-MM-DD` form. Invalid date strings
are not reported as future dates.

### Counts reflect REDCap's row structure

A REDCap record can produce multiple API rows because of events and repeating
instances. Project `record_count` counts distinct records, `raw_row_count`
counts exported rows, field `record_count` counts applicable rows, and event
`row_count` counts exported rows per event. Choose the measure that matches the
question being asked.

### The report depends on current REDCap structure

Descriptive fields are removed before analysis because they do not store record
values. Structural metadata retrieval failures are reported as errors rather
than silently converted to empty project structure. Results reflect the API
records and metadata available when the report is built, including the token's
project access and export configuration.
