---
title: "Understanding REDCap Quality Reports"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(REDCapExploreR)
```

`build_quality_report()` retrieves REDCap records and project structure, applies
general data quality checks, and organizes the results for review. The report is
intended to help data managers and analysts identify records, fields, and
metadata that warrant attention. It does not modify the REDCap project.

This article explains how to build and navigate a report. All evaluated
examples use `mock_quality_report`, a synthetic report included with the
package, so they do not require REDCap credentials.

## How to Use

### Build a report

Store the REDCap API URI and token outside version-controlled code, such as in
environment variables, and pass them to `build_quality_report()`:

```{r build-report, eval=FALSE}
report <- build_quality_report(
  redcap_uri = Sys.getenv("REDCAP_URI"),
  token = Sys.getenv("REDCAP_TOKEN")
)
```

`checks` controls which findings are generated. Summaries and interpreted
metadata are still returned, while `finding_count` values in the project and
record summaries reflect the selected check groups. `mock_quality_report` was
built with all default check groups and thresholds.

The default call runs all five check groups: `missingness`, `metadata`,
`outliers`, `operational`, and `consistency`. Use `checks` to run a focused
review:

```{r focused-report, eval=FALSE}
report <- build_quality_report(
  redcap_uri = Sys.getenv("REDCAP_URI"),
  token = Sys.getenv("REDCAP_TOKEN"),
  checks = c("missingness", "metadata"),
  progress = "none"
)
```

Two arguments control heuristic thresholds:

- `sparse_threshold` is the applicable-row missing rate at which a field is
  reported as unexpectedly sparse. It defaults to `0.95`.
- `outlier_iqr_multiplier` controls the distance from the first and third
  quartiles used by the IQR outlier check. It defaults to `3`.

`progress = "auto"` displays progress in interactive sessions. Use `"show"` to
force it or `"none"` to suppress it.

### Start with the report overview

Printing the report gives a short project-level overview. The object itself is
a list with `findings`, `summaries`, and `metadata` elements.

```{r report-overview}
mock_quality_report

names(mock_quality_report)
names(mock_quality_report$summaries)
names(mock_quality_report$metadata)
```

Begin with the project summary and a count of findings by check and issue. This
shows the size and structure of the project alongside the types of review items
that were found.

```{r report-triage}
mock_quality_report$summaries$project

mock_quality_report$findings |>
  dplyr::count(check, issue, severity, sort = TRUE)
```

### Review actionable findings

Use `scope` to distinguish findings tied to a specific record and field from
field-level metadata findings. The context columns remain available for
longitudinal and repeating projects.

```{r record-findings}
record_findings <- mock_quality_report$findings |>
  dplyr::filter(scope == "record_field") |>
  dplyr::select(
    finding_id,
    severity,
    issue,
    record_id,
    event_name,
    repeat_instrument,
    repeat_instance,
    field_name,
    value,
    expected
  )

record_findings
```

Filter by record, issue, form, or severity to create a review queue. Severity is
a package-provided prioritization aid, not a clinical or regulatory
classification.

```{r review-one-record}
mock_quality_report$findings |>
  dplyr::filter(record_id == "C002") |>
  dplyr::select(
    finding_id,
    severity,
    issue,
    form_name,
    field_name,
    value,
    expected,
    message
  )
```

Join findings to interpreted field metadata when labels and field types are
useful during review:

```{r enrich-findings}
field_reference <- mock_quality_report$metadata$fields |>
  dplyr::select(field_name, field_label, field_type, required_field)

mock_quality_report$findings |>
  dplyr::filter(!is.na(field_name)) |>
  dplyr::inner_join(field_reference, by = "field_name") |>
  dplyr::select(
    finding_id,
    issue,
    record_id,
    field_name,
    field_label,
    field_type,
    required_field
  )
```

## Key Elements

### Findings

`findings` is a standardized tibble with one row per detected issue. A report
with no findings returns an empty tibble with the same columns.

| Column | Meaning |
|---|---|
| `finding_id` | Sequential identifier assigned within the report. |
| `check` | Check group that produced the finding. |
| `issue` | Stable issue name suitable for filtering and counting. |
| `severity` | Package-assigned `warning` or `info` review priority. |
| `scope` | `record_field` for record-level findings or `field` for metadata and aggregate findings. |
| `record_id` | REDCap record identifier when the finding is record-specific. |
| `form_name` | REDCap instrument associated with the finding. |
| `event_name` | Unique REDCap event name when available. |
| `repeat_instrument` | Repeating instrument name when available. |
| `repeat_instance` | Repeating instance number when available. |
| `field_name` | REDCap field associated with the finding. |
| `value` | Observed value or aggregate value that triggered the finding. |
| `expected` | Concise description of the expected state or range. |
| `message` | Human-readable explanation of the finding. |

The following checks are currently performed.

| Check | Issue | What it identifies | Severity and scope |
|---|---|---|---|
| Missingness | `required_field_missing` | A required field is blank on a row where the form and branching logic are applicable and completion status permits missingness assessment. | Warning, `record_field` |
| Missingness | `unexpected_sparse_field` | A field's missing rate is at or above `sparse_threshold` among its applicable rows. | Info, field |
| Metadata | `duplicate_field_label` | The same nonblank field label is used by multiple field names. | Info, field |
| Metadata | `missing_field_label` | A field has no label. | Info, field |
| Metadata | `missing_choice_definition` | A radio, dropdown, or checkbox field has no explicit choice definition. | Warning, field |
| Metadata | `orphaned_branching_reference` | Branching logic references a field not present in project metadata. | Warning, field |
| Metadata | `high_risk_free_text` | A text or notes field name or label contains terms associated with notes, comments, other details, or similar free text. | Info, field |
| Outliers | `outside_validation_range` | A numeric value is below or above a REDCap metadata validation bound. | Warning, `record_field` |
| Outliers | `numeric_iqr_outlier` | A numeric, non-choice field falls outside the configured IQR heuristic. At least four observed values and a nonzero IQR are required. | Info, `record_field` |
| Outliers | `future_date` | A field with date validation contains a `YYYY-MM-DD` value after the report date. | Info, `record_field` |
| Operational | `incomplete_form_status` | An applicable form status is missing or is not Complete. | Info, `record_field` |
| Consistency | `checkbox_none_with_other` | A checkbox option labeled as none, no, or not applicable is selected with another option. | Warning, `record_field` |

Choice fields are excluded from IQR outlier detection even when their codes are
numeric. Yes/No and True/False fields have implicit REDCap choices, so they do
not require an explicit choice definition.

### Summaries

`summaries` provides five views of the normalized project and report findings.

#### Project

`summaries$project` is a one-row overview:

- `record_count`: number of distinct record identifiers.
- `raw_row_count`: number of rows returned by the records API, including event
  and repeating rows.
- `field_count`, `form_count`, and `event_count`: project structure counts.
- `repeating_enabled`: whether repeating configuration was returned by REDCap.
- `missing_rate`: weighted missing fraction across applicable field-row cells.
- `finding_count`: total number of findings from the selected check groups.

```{r project-summary}
mock_quality_report$summaries$project
```

#### Forms and fields

`summaries$forms` reports field counts, required-field counts, and weighted
missing rates by form. `summaries$fields` gives the most detailed missingness
profile:

- `record_count` is the number of applicable rows assessed for the field. It is
  not necessarily the number of distinct records because longitudinal and
  repeating projects can contribute multiple rows per record.
- `missing_count` and `observed_count` partition applicable rows.
- `missing_rate` is `missing_count / record_count`, or `NA` when no rows are
  applicable.
- `distinct_count` counts unique nonmissing values.

```{r field-summaries}
mock_quality_report$summaries$forms

mock_quality_report$summaries$fields |>
  dplyr::arrange(dplyr::desc(missing_rate), field_name) |>
  dplyr::select(
    field_name,
    form_name,
    required_field,
    record_count,
    missing_count,
    missing_rate,
    distinct_count
  )
```

Project and form missing rates are weighted across applicable field-row cells.
They are not simple averages of field-level percentages.

#### Records and events

`summaries$records` contains one row per record. `missing_field_count` sums
missing applicable field values across that record's events and repeat
instances, while `finding_count` counts record-specific findings. Field-level
metadata findings do not belong to an individual record.

`summaries$events` counts raw API rows by unique event name. It is empty for
nonlongitudinal projects.

```{r record-event-summaries}
mock_quality_report$summaries$records |>
  dplyr::arrange(dplyr::desc(finding_count), record_id)

mock_quality_report$summaries$events
```

### Metadata

`metadata` preserves and interprets project structure used by the checks.

| Element | Contents |
|---|---|
| `fields` | Standardized REDCap field metadata used by the report. |
| `forms` | One row per form with field, required-field, and choice-field counts. |
| `events` | Standardized longitudinal event metadata. |
| `event_instruments` | Standardized event-to-form mapping. |
| `instruments` | Instrument names and labels returned by REDCap. |
| `repeating_instruments` | Repeating event and instrument configuration returned by REDCap, with normalized join columns. |
| `choices` | Parsed explicit choices and implicit Yes/No and True/False choices. |
| `validation` | Fields with validation types or minimum/maximum bounds. |
| `branching` | Fields with branching logic and parsed referenced field names. |
| `record_id_field` | Field name used as the project record identifier. |
| `source` | Source of the normalized project; API reports use `"api"`. |

These tables are useful for interpreting findings and for project-level review
that is not represented by a finding. For example, inspect repeating structure
and validation rules directly:

```{r metadata-examples}
mock_quality_report$metadata$repeating_instruments

mock_quality_report$metadata$validation

mock_quality_report$metadata$branching
```

## Core Assumptions

### Findings are review prompts

The checks are broad heuristics intended to identify items worth reviewing.
They do not establish clinical correctness, protocol compliance, or regulatory
compliance. An `info` finding can still be important, and a report with no
findings does not prove that the data are correct.

`finding_id` is assigned when a report is built and should not be treated as a
persistent identifier across runs. Use issue and REDCap context columns when
tracking findings externally.

### Missingness is assessed only where a field is applicable

`NA` and blank or whitespace-only values are treated as missing. For checkbox
fields, one or more checked options make the field observed; a set of unchecked
exports is treated as missing when the field is assessed.

Applicability combines several pieces of project structure:

1. Longitudinal forms are limited to events in the event-to-form mapping.
2. Repeating instruments are identified from REDCap repeating configuration,
   not inferred solely from rows currently present in the data export.
3. Repeating rows are matched to their configured instrument so structural
   blanks from other repeating instruments are not counted as missing.
4. Branching logic limits assessment to rows where the field should be shown.
5. Missingness is assessed only when `<form_name>_complete` is Unverified (`1`)
   or Complete (`2`). Incomplete (`0`) forms are handled by the operational
   check instead.

This design avoids treating structural blanks as missing data. It also means a
field can have `record_count = 0` and `missing_rate = NA` when no exported rows
are eligible for assessment.

### Branching logic support is intentionally bounded

The branching evaluator supports ordinary field comparisons, checkbox choice
references, parentheses, and `and`/`or` combinations. A valid expression that
evaluates to `NA` is treated as not shown for that row.

REDCap functions, smart variables, and other expressions outside the supported
subset are treated as applicable rather than used to hide a potentially
required value. The metadata check separately reports references to fields that
are absent from project metadata. Complex branching findings should therefore
be reviewed against REDCap itself.

### Outlier checks depend on metadata and sample size

Validation-range findings depend on numeric minimum and maximum values defined
in REDCap metadata. IQR findings are distribution-based and require at least
four observed values and a nonzero IQR. Their usefulness depends on the field's
distribution and the selected multiplier.

Future-date checks apply only to fields whose REDCap validation type contains
`date`, and values are interpreted in `YYYY-MM-DD` form. Invalid date strings
are not reported as future dates.

### Completion status drives two different questions

Missingness asks whether expected values are present on forms sufficiently
complete to assess. The operational check asks whether accessible forms have a
Complete status. For that reason, a form marked Incomplete can produce an
operational finding without contributing required-field missingness findings.

### Counts reflect REDCap's row structure

A REDCap record can produce multiple API rows because of events and repeating
instances. Project `record_count` counts distinct records, `raw_row_count`
counts exported rows, field `record_count` counts applicable rows, and event
`row_count` counts exported rows per event. Choose the measure that matches the
question being asked.

### The report depends on current REDCap structure

Descriptive fields are removed before analysis because they do not store record
values. Structural metadata retrieval failures are reported as errors rather
than silently converted to empty project structure. Results reflect the API
records and metadata available when the report is built, including the token's
project access and export configuration.
